---
title: "Community annalisys in Vegan"
author: "Julian Avila-Jimenez"
date: "28/9/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)
library(vegan)
#remotes::install_github("gavinsimpson/ggvegan")
library(ggvegan)
library(tidyverse)
```


## Community annalisys in the R package (*vegan*)[https://github.com/vegandevs/vegan] 
```{r Data}
#load mite data (use ?mite to get more information)
data(mite) #community data on mites collected from 70 soil cores
data(mite.env) #Environmental data collevctes from cores locations 
data(mite.pcnm) #Principal components of neighbor matrices
```

### Unconstrained ordination example

#### PCA
A (PCA)[https://en.wikipedia.org/wiki/Principal_component_analysis] is a linear combination of the explanatory variables, is a method to reduce dimensionality.
```{r PCA}
pca<- rda(mite.env[,1:2]) #pca only on the numerical variables substrate density and water content
summary(pca)
autoplot(pca, arrows = T)
```

Inertia means the amount of variance that is explained by the environmental variables. In the *summary* The eigen values of each PC represents the amount of inertia of that particular PC. 

